Aquifer depletion exacerbates agricultural drought losses in the US High Plains

### 2023-10-31 Writing/manuscript.tex
clean up the codes

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{lineno}
\linenumbers
\usepackage{float}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Aquifer depletion exacerbates agricultural drought losses in the US High Plains},
  pdfauthor={Taro Mieno; Timothy Foster; Shunkei Kakimoto; Nicholas Brozovic},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Aquifer depletion exacerbates agricultural drought losses in the US High Plains}
\author{Taro Mieno\footnote{Department of Agricultural Economics, University of Nebraska Lincoln, \href{mailto:tmieno2@unl.edu}{\nolinkurl{tmieno2@unl.edu}}} \and Timothy Foster\footnote{Department of WWW, University of Manchester, \href{mailto:timothy.foster@manchester.ac.uk}{\nolinkurl{timothy.foster@manchester.ac.uk}}} \and Shunkei Kakimoto\footnote{Department of Applied Economics, University of Minnesota, \href{mailto:kakim002@umn.edu}{\nolinkurl{kakim002@umn.edu}}} \and Nicholas Brozovic\footnote{Department of Agricultural Economics, University of Nebraska Lincoln, \href{mailto:nbrozovic@nebraska.edu}{\nolinkurl{nbrozovic@nebraska.edu}}}}
\date{}

@@

Despite widespread alarm about the risks that aquifer depletion poses to crop productivity and food security, empirical evidence is limited about how reductions in groundwater availability will alter farmers capacity to adapt successfully to drought and rainfall variability. Most empirical studies that have assessed impacts of drought on agricultural productivity have focused primarily on rainfed production areas due to the greater exposure of rainfed agriculture to drought-related shocks \citep{schlenker2009nonlinear, lobell2014greater, schlenker2010robust, zhou2020connections, borgomeo2020impact}. Existing studies also demonstrate the importance of existing irrigated areas or future irrigation intensification and expansion to mitigate negative impacts of drought \citep{kuwayama2019estimating, zipper2016drought, zhu2022untangling, zhu2022warming, lu2020mapping, davis2019sensitivity, li2018changes}. In contrast, most research assessing impacts of aquifer depletion on resilience of irrigated farmland to drought has relied on simulation modeling \citep{foster2015well, cotterman2018groundwater, kahil2015modeling, yoon2021coupled, rad2020mod}, with more limited work that attempts to empirically evaluate change in drought-related production losses and risks on irrigated lands as a function of changing aquifer storage \citep{jain2021groundwater, suter2021depletion}.

Two pathways exist by which aquifer depletion will impact farmers' ability to access and use groundwater as a buffer against drought and rainfall variability \citep{foster2015analysis}. First, depletion increases the energy requirements and costs of pumping water, potentially making it less economically viable for farmers to satisfy crop water needs \citep{mieno2017price, bhattarai2021impact}. Second, depletion also reduces the thickness of an aquifer, lowering the aquifer's transmissivity (i.e., the potential rate of water supply to wells). This latter effect reduces the rate at which water can be pumped and applied to a field as irrigation \citep{konikow2005groundwater, foster2014modeling, hrozencik2017heterogeneous}. If the reduction in well yields is substantial enough - for example as would occur when a large proportion of aquifer thickness is depleted \citep{hecox2002calculation, korus2020depletion} - then a farmer may be unable to extract sufficient water to meet crop water demands, in particular during periods of limited rainfall and in the peak of the growing season when crops are most sensitive to water deficits \citep{foster2015well, rouhi2020downside}. The combined effect of declining well yields and increasing pumping costs could force farmers to reduce irrigation depths, resulting in greater risk of drought-related crop stress. Alternatively, or in addition, a farmer may be forced to limit irrigated area to ensure crop water demands can be met, leading to reductions in total crop production \citep{foster2014modeling, rad2020effects}. Understanding how these two responses vary as a function of aquifer conditions is crucial to guide groundwater management planning decisions, in particular given projected increases in frequency and intensity of droughts under climate change \citep{ukkola2020robust, chiang2021evidence, cook2020twenty}.

@@

Figure \ref{fig:yield-response} shows the estimated crop yield response to water deficit for rainfed production and groundwater-irrigated production at different levels of aquifer thickness categories for corn (panel A) and soybean (panel B)\footnote{Confidence intervals were not presented to avoid over-crowding the figures. Please see Appendix \ref{reg-conf} for estimated yield response curves with 95 percent confidence interval.}. The histogram of annual water deficit is presented at the top for each crop. In wetter years with low crop irrigation requirements (i.e., negative values of water deficit), rainfed and irrigated yields are very similar for both corn and soybean. As drought severity increases (i.e., as water deficit increases), rainfed yields of both crops decline significantly whereas irrigated yields are increasing or stable as deficit increases. This result reflects the positive benefits irrigation provides both as a buffer against precipitation variability, and as a means of raising overall yields through increased evapotranspiration and associated biomass accumulation, reductions of heat-related stressors, and other factors \citep{zhu2022untangling, li2020quantifying}.

We now focus on comparing yield response to water deficit among the three aquifer thickness categories (all irrigated). First, for water deficits above around 400mm, yield changes become negative for counties with the lowest levels of aquifer thickness (i.e., 1st quantile in Figure \ref{fig:yield-response}). This likely reflects the restrictions that lower aquifer thickness and associated reductions in well yields place on farmers ability to fully meet crop water requirements during periods of more severe or extended water deficits \citetext{\citealp[\citet{foster2014modeling}]{rad2020effects}; \citealp{hrozencik2017heterogeneous}}. In contrast, yield continues to increase gradually as water deficit increases beyond 400 mm for counties with the highest level of aquifer thickness.

@@

At low water deficits (below 200mm), Figure \ref{fig:yield-dif} illustrates that counties with the highest levels of aquifer thickness (i.e., 3rd quantile in Figure \ref{fig:yield-response} experience statistically lower yields for corn and soybean than counties with the lowest level of aquifer thickness (1st quantile). In case of soybean, irrigated yields for the lowest aquifer thickness category converge to levels comparable to those obtained from rainfed production. These counter-intuitive results reflect a combination of two factors. First, in years with small water deficits there is less need for irrigation and hence irrigated and rainfed yield differences are reduced. Second, years with small or even negative water deficits are associated with wetter conditions during the growing season, which can negatively impact yield if waterlogging occurs leading to greater risks of physical crop damage (e.g., lodging of grains), delays to planting/harvesting, nutrient leaching, and restricted root growth in waterlogged soils \citep{li2019excessive}. Declines in irrigated yields could be indicative of over-irrigation by farmers, which has been shown to occur in the HPA region in wet years \citep{foster2019assessing, gibson2017case, gibson2019benchmarking}, exacerbating these risks. However, the small declines in irrigated yields at low or negative levels of water deficit may also reflect the fact that wetter years are typically cooler, reducing positive heat effects on crop growth and yield formation.

@@

{\centering \includegraphics[width=6.5in,]{Figures/g_yield_intensive}

@@

\caption{The impact of water deficit and aquifer thickness on rainfed and irrigated per-area yields of corn and soybean in US High Plains. Irrigated yield responses to water deficits are dissagregated by aquifer thickness level, considering three quantile groups that capture the range of aquifer thickness values in our sample: (i) 9 m to 30.1 m; (ii) 30.1 m to 67.2 m; and (iii) 67.2 m to 215.8 m. The histograms on the top of each panel show the distribution of county-level water deficit values included in regressions for each crop.}\label{fig:yield-response}
\end{figure}

@@

{\centering \includegraphics[width=6in,]{Figures/g_ir_yield_dif}

@@

Figure \ref{fig:ir-share} shows the estimated relationship between aquifer thickness and the share of production area that is irrigated for corn and soybean after accounting for spatio-temporal variations in weather conditions and drought risk exposure across the aquifer. Figure \ref{fig:ir-share} demonstrates that aquifer thickness has meaningfully large impacts on the share of irrigated production for both corn and soybean. As aquifer thickness declines, the share of irrigated acres for corn decreases from around 0.89 (89\%) for an average county-level aquifer thickness of 130m to 0.72 (72\%) for an average county-level aquifer thickness of 10m (See Appendix \ref{test-dif-share} for formal statistical testings). Reductions in the share of production area under irrigation is also evident for soybean, although the effect is less pronounced than for corn, potentially due to soybean's lower water requirements and susceptibility to drought relative to corn \citep{zipper2016drought, ruess2022irrigation}

@@

{\centering \includegraphics[width=5in,]{Figures/g_share}

@@

\caption{The estimated impact of aquifer thickness on the share of agricultural production under irrigation for corn and soybean producing counties in the US High Plains (Note: shaded area represents 95 percent confidence interval)}\label{fig:ir-share}
\end{figure}

@@

The overall influence of aquifer depletion on drought-related production losses is a function of both per-area irrigated yield and irrigated area changes highlighted in Sections \ref{impact-intensive} and \ref{impact-share}. Combining both responses, Figure \ref{fig:tot-impact} shows the relationship between water deficit and average per-area production (weighted-average of irrigated and rainfed yields, where the weights are production area shares) for corn and soybean at four values of aquifer thickness: 10, 50, 90, and 130 meters. While similar trends are observed as in Figure \ref{fig:yield-response}, the difference in production loss caused by high water deficits among the aquifer thickness levels are much more pronounced due to lower estimated irrigated area share for smaller aquifer thickness values. For example, for an aquifer thickness of 10 m, reductions in average yields occur for smaller water deficits than are required to trigger reductions in per-area irrigated yields alone. Similarly, the overall magnitude of yield reduction is greater for average yields than for per-area irrigated yields. This is because low aquifer thickness values have a greater estimated share of crop area under rainfed production, which magnifies average yield declines with increasing water deficits.

Figure \ref{fig:dif-tot-impact} shows the difference in average per-area productivity for 50 m, 90 m, and 130 m aquifer thickness levels relative to an aquifer thickness of 10 m for corn and soybean. For corn, aquifer thickness does not have a significant impact on productivity when water deficits are below about 350mm. However, beyond this threshold, higher levels of water deficit result in noticeable and statistically significant reductions to average corn production. This reflects the combination of lower irrigated yields under drought conditions, and, more importantly, reduced overall productivity due to smaller share of land under irrigation in areas and years where aquifer thickness is lower. Similar effects are also observed for soybean, although reductions in average production only occur at a higher deficit threshold (greater than around 600mm) and are smaller in overall magnitude. For example, for a seasonal water deficit of 950mm (average water deficit in 2012, the most extreme drought in our historical record), average per-area corn and soybean productivity are about 25\% and 15\% lower, respectively, when differentiating between areas with the 130m and 10m aquifer thicknesses.

Figure \ref{fig:tot-impact} and Figure \ref{fig:dif-tot-impact} also demonstrate that impact of aquifer thickness on average per-area production is highly non-linear at high seasonal water deficits. At a water deficit of 950mm, for example, reducing from 130m aquifer thickness to 90m (decline of 40m) would result in only 0.14 tonne/ha of production loss for corn. This negligible difference in average yield from the decline is due to the insensitivity of the share of irrigated acres to a decline in aquifer thickness when the initial aquifer thickness level is high as can be seen in Figure \ref{fig:ir-share}. Reducing from 90m aquifer thickness to 50m (decline of 40m) would result in 0.94 tonne/ha of production loss for corn, with a further drop from 50m aquifer thickness to 10m (decline of 40m) leading in 1.44 tonne/ha of production loss for corn. Similar but less pronounced effects are also observed for soybean. This non-linear impact of aquifer thickness is due to the influence of aquifer thickness on the share of irrigated corn acres (Figure \ref{fig:ir-share}), which implicitly reflects how aquifer thickness alters well yields and, in turn, farmers ability to effectively and reliably buffer crops against water deficits during the growing season.

@@

{\centering \includegraphics[width=6in,]{Figures/g_total_impact}

@@

\caption{Average productivity of corn and soybean in the US High Plains for different levels of water deficit and aquifer thickness, considering both per-area yield responses and farmers choices about irrigated area shares in response to different water deficit and aquifer conditions. See Appendix A for estimated average yield response curves with 95 percent confidence intervals displayed, which are omitted here for visual clarity.}\label{fig:tot-impact}
\end{figure}

@@

{\centering \includegraphics[width=6in,]{Figures/g_avg_yield_dif}

@@

\caption{Difference in average yield of corn and soybean for aquifer thickness levels of 50 m, 90 m, and 130 m, relative to a baseline aquifer thickness of 10 m. Shaded areas denote 95 percent confidence interval, which when greater than zero indicate average yields for the given aquifer thickness level are statistically significantly greater than for an aquifer thickness of 10 m at the 5 percent level of significance.}\label{fig:dif-tot-impact}
\end{figure}

@@

Second, our analysis considered the effects of aquifer thickness on farmers decisions to switch from irrigated to rainfed production, but did not assess other potential responses to aquifer depletion such as changing crop types or irrigation management practices. Switching to more drought resistant crops or varieties may enable farmers to maintain irrigated production at lower levels of aquifer thickness, and has been shown to be an important response to increasing agricultural water scarcity and abstraction policies in other studies \citep{bhattarai2021impact, deines2019quantifying, manning2017producer}. Improvements to irrigation scheduling practices and technologies, such as have been implemented as part of the SD6-LEMA (Local Enhanced Management Area) in Kansas \citep{deines2019quantifying, glose2022quantifying} and in parts of Texas \citep{mrad2020peak} and which are proposed as part of California's Sustainable Groundwater Management Act (SGMA) \citep{berbel2019droughts, lubell2020sustainable}, could also help to improve the efficiency of irrigation and, hence, the ability to more effectively meet crop water demands with lower well capacities. Such interventions may reduce - but not eliminate - increases in drought risk caused by lower or declining aquifer thickness, and could also help to mitigate against potential yield losses caused by over-use of water in wetter years via raising awareness of efficiency saving potentials \citep{foster2019assessing} and the importance of conservation \citep{marston2022}. However, risks also exist that irrigation efficiency improvements will increase net extraction, ultimately causing a rebound effect that exacerbates future aquifer depletion and drought risks \citep{grafton2018paradox, perez2021agricultural}. Future research should explore potential strategies that farmers could adopt to lessen the combined impacts of drought and aquifer depletion, alongside efforts to curb unsustainable groundwater extraction and climate change.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{lineno}
\linenumbers
\usepackage{float}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{authblk}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Aquifer depletion exacerbates agricultural drought losses in the US High Plains},
  pdfauthor={Taro Mieno; Timothy Foster; Shunkei Kakimoto; Nicholas Brozovic},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Aquifer depletion exacerbates agricultural drought losses in the US High Plains}

\author[1]{Taro Mieno}
\author[2]{Timothy Foster}
\author[3]{Shunkei Kakimoto}
\author[4]{Nicholas Brozovi\'{c}}
\affil[1]{Department of Agricultural Economics, University of Nebraska Lincoln}
\affil[2]{School of Mechanical, Aerospace and Civil Engineering, University of Manchester}
\affil[3]{Department of Applied Economics, University of Minnesota}
\affil[4]{Daugherty Water for Food Global Institute at the University of Nebraska}
\affil[4]{Department of Agricultural Economics, University of Nebraska Lincoln}

\date{}

@@

Despite widespread alarm about the risks that aquifer depletion poses to crop productivity and food security, empirical evidence is limited about how reductions in groundwater availability will alter farmers' capacity to adapt successfully to drought and rainfall variability. Most empirical studies that have assessed impacts of drought on agricultural productivity have focused primarily on rainfed production areas due to the greater exposure of rainfed agriculture to drought-related shocks \citep{schlenker2009nonlinear, lobell2014greater, schlenker2010robust, zhou2020connections, borgomeo2020impact}. Existing studies also demonstrate the importance of existing irrigated areas or future irrigation intensification and expansion to mitigate negative impacts of drought \citep{kuwayama2019estimating, zipper2016drought, zhu2022untangling, zhu2022warming, lu2020mapping, davis2019sensitivity, li2018changes}. In contrast, most research assessing impacts of aquifer depletion on resilience of irrigated farmland to drought has relied on simulation modeling \citep{foster2015well, cotterman2018groundwater, kahil2015modeling, yoon2021coupled, rad2020mod}, with more limited work that attempts to empirically evaluate change in drought-related production losses and risks on irrigated lands as a function of changing aquifer storage \citep{jain2021groundwater, suter2021depletion}.

Two major pathways exist by which aquifer depletion will impact farmers' ability to access and use groundwater as a buffer against drought and rainfall variability \citep{foster2015analysis}\footnote{Another pathway is through groundwater use regulation in response to low or declining aquifer thickness. However, it is unlikely that this effects is captured in our analysis. Please see Appendix \ref{pathway-reg}}. First, depletion increases the energy requirements and costs of pumping water, potentially making it less economically viable for farmers to satisfy crop water needs \citep{mieno2017price, bhattarai2021impact}. Second, depletion also reduces the thickness of an aquifer, lowering the aquifer's transmissivity (i.e., the potential rate of water supply to wells). This latter effect reduces the rate at which water can be pumped and applied to a field as irrigation \citep{konikow2005groundwater, foster2014modeling, hrozencik2017heterogeneous}. If the reduction in well yields is substantial enough - for example as would occur when a large proportion of aquifer thickness is depleted \citep{hecox2002calculation, korus2020depletion} - then a farmer may be unable to extract sufficient water to meet crop water demands, in particular during periods of limited rainfall and in the peak of the growing season when crops are most sensitive to water deficits \citep{foster2015well,rouhi2020downside}. The combined effect of declining well yields and increasing pumping costs could force farmers to reduce irrigation depths, resulting in greater risk of drought-related crop stress. Alternatively, or in addition, a farmer may be forced to limit irrigated area to ensure crop water demands can be met, leading to reductions in total crop production \citep{foster2014modeling, rad2020effects}. Understanding how these two responses vary as a function of aquifer conditions is crucial to guide groundwater management planning decisions, in particular given projected increases in frequency and intensity of droughts under climate change \citep{ukkola2020robust, chiang2021evidence, cook2020twenty}.

@@

Figure \ref{fig:yield-response} shows the estimated crop yield response to water deficit for rainfed production and groundwater-irrigated production at different levels of aquifer thickness categories for corn (panel A) and soybean (panel B)\footnote{Confidence intervals were not presented to avoid over-crowding the figures. Please see Appendix \ref{reg-conf} for estimated yield response curves with 95 percent confidence interval.}. The histogram of annual water deficit is presented at the top for each crop. In wetter years with low crop irrigation requirements (i.e., negative values of water deficit), rainfed and irrigated yields are very similar for both corn and soybean. As drought severity increases (i.e., as water deficit increases), rainfed yields of both crops decline significantly whereas irrigated yields are increasing or stable as deficit increases (see Figures \ref{fig:irrigated-yield-ind-corn} and \ref{fig:irrigated-yield-ind-soy} for the 95\% confidence intervals). This result reflects the positive benefits irrigation provides both as a buffer against precipitation variability, and as a means of raising overall yields through increased evapotranspiration and associated biomass accumulation, reductions of heat-related stressors, and other factors \citep{zhu2022untangling, li2020quantifying}.

We now focus on comparing yield response to water deficit among the three aquifer thickness categories (all irrigated). First, for water deficits above around 400mm, yield changes become negative for counties with the lowest levels of aquifer thickness (i.e., 1st quantile in Figure \ref{fig:yield-response}). This likely reflects the restrictions that lower aquifer thickness and associated reductions in well yields place on farmers' ability to fully meet crop water requirements during periods of more severe or extended water deficits \citetext{\citealp[\citet{foster2014modeling}]{rad2020effects}; \citealp{hrozencik2017heterogeneous}}. In contrast, yield continues to increase gradually as water deficit increases beyond 400 mm for counties with the highest level of aquifer thickness.

@@

% At low water deficits (below 200mm), Figure \ref{fig:yield-dif} illustrates that counties with the highest levels of aquifer thickness (i.e., 3rd quantile in Figure \ref{fig:yield-response}) experience statistically lower yields for corn and soybean than counties with the lowest level of aquifer thickness (1st quantile). In case of soybean, irrigated yields for the lowest aquifer thickness category converge to levels comparable to those obtained from rainfed production. These counter-intuitive results reflect a combination of two factors. First, in years with small water deficits there is less need for irrigation and hence irrigated and rainfed yield differences are reduced. Second, years with small or even negative water deficits are associated with wetter conditions during the growing season, which can negatively impact yield if waterlogging occurs leading to greater risks of physical crop damage (e.g., lodging of grains), delays to planting/harvesting, nutrient leaching, and restricted root growth in waterlogged soils \citep{li2019excessive}. Declines in irrigated yields could be indicative of over-irrigation by farmers, which has been shown to occur in the HPA region in wet years \citep{foster2019assessing, gibson2017case, gibson2019benchmarking}, exacerbating these risks. However, the small declines in irrigated yields at low or negative levels of water deficit may also reflect the fact that wetter years are typically cooler, reducing positive heat effects on crop growth and yield formation.

@@

{\centering \includegraphics[width=6.5in,height=500px,]{../../Figures/g_yield_intensive}

@@

\caption{The impact of water deficit and aquifer thickness on rainfed and irrigated per-area yields of corn and soybean in US High Plains. Irrigated yield responses to water deficits are dissagregated by aquifer thickness level, considering three quantile groups that capture the range of aquifer thickness values in our sample: (i) 9 m to 30.1 m; (ii) 30.1 m to 67.2 m; and (iii) 67.2 m to 215.8 m. The histograms on the top of each panel show the distribution of county-level annual water deficit values included in regressions for each crop. Total number of observations are $8,773$ and $5,977$ for corn and soybean regressions, respectively.}\label{fig:yield-response}
\end{figure}

@@

{\centering \includegraphics[width=6in,height=500px,]{../../Figures/g_ir_yield_dif}

@@

Figure \ref{fig:ir-share} shows the estimated relationship between aquifer thickness and the share of production area that is irrigated for corn and soybean after accounting for spatio-temporal variations in weather conditions and drought risk exposure across the aquifer. Figure \ref{fig:ir-share} demonstrates that aquifer thickness has meaningfully large impacts on the share of irrigated production for both corn and soybean. As aquifer thickness declines, the share of irrigated acres for corn decreases from around 0.90 (90\%) for an average county-level aquifer thickness of 100 m to 0.72 (72\%) for an average county-level aquifer thickness of 10 m (See Appendix \ref{test-dif-share} for formal statistical testings). Reductions in the share of production area under irrigation is also evident for soybean.

@@

{\centering \includegraphics[width=6in,height=500px,]{../../Figures/g_share}

@@

\caption{The estimated impact of aquifer thickness on the share of agricultural production under irrigation for corn and soybean producing counties in the US High Plains (Note: shaded area represents 95\% confidence interval). The histograms on the top of each panel show the distribution of county-level aquifer thickness values included for each crop. Total number of observations are $2,523$ and $1,558$ for corn and soybean regressions, respectively.}\label{fig:ir-share}
\end{figure}

@@

The overall influence of aquifer depletion on drought-related production losses is a function of both per-area irrigated yield and irrigated area changes highlighted in Sections \ref{impact-intensive} and \ref{impact-share}. Combining both responses, Figure \ref{fig:tot-impact} shows the relationship between water deficit and average per-area production (weighted-average of irrigated and rainfed yields, where the weights are production area shares) for corn and soybean at four values of aquifer thickness: 10, 40, 70, and 100 meters. While similar trends are observed as in Figure \ref{fig:yield-response}, the difference in production loss caused by high water deficits among the aquifer thickness levels are much more pronounced due to lower estimated irrigated area share for smaller aquifer thickness values. For example, for an aquifer thickness of 10 m, reductions in average yields occur for smaller water deficits than are required to trigger reductions in per-area irrigated yields alone. Similarly, the overall magnitude of yield reduction is greater for average yields than for per-area irrigated yields. This is because low aquifer thickness values have a greater estimated share of crop area under rainfed production, which magnifies average yield declines with increasing water deficits.

Figure \ref{fig:dif-tot-impact} shows the difference in average per-area productivity for 40 m, 70 m, and 100 m aquifer thickness levels relative to an aquifer thickness of 10 m for corn and soybean. For corn, aquifer thickness does not have a significant impact on productivity when water deficits are below about 400 mm. However, beyond this threshold, higher levels of water deficit result in noticeable and statistically significant reductions to average corn production. This reflects the combination of lower irrigated yields under drought conditions, and, more importantly, reduced overall productivity due to smaller share of land under irrigation in areas where aquifer thickness is lower. Similar effects are also observed for soybean, although reductions in average production only occur at a higher deficit threshold (greater than around 600 mm) and are smaller in overall magnitude. For example, for a seasonal water deficit of 950 mm (average water deficit in 2012, the most extreme drought in our historical record), average per-area corn and soybean productivity are about 25\% and 15\% lower, respectively, when differentiating between areas with the 100 m and 10 m aquifer thicknesses.

Figure \ref{fig:tot-impact} and Figure \ref{fig:dif-tot-impact} also demonstrate that impact of aquifer thickness on average per-area production is highly non-linear at high seasonal water deficits. At a water deficit of 950 mm, for example, reducing from 100 m aquifer thickness to 70 m (decline of 30 m) would result in only 0.09 tonne/ha of production loss for corn. This negligible difference in average yield from the decline is due to the insensitivity of the share of irrigated acres to a decline in aquifer thickness when the initial aquifer thickness level is high as can be seen in Figure \ref{fig:ir-share}. Reducing from 70 m aquifer thickness to 40 m (decline of 30 m) would result in 0.95 tonne/ha of production loss for corn, with a further drop from 40 m aquifer thickness to 10 m (decline of 30 m) leading to 1.57 tonne/ha of production loss for corn. Similar but less pronounced effects are also observed for soybean. This non-linear impact of aquifer thickness is due to the influence of aquifer thickness on the share of irrigated corn acres (Figure \ref{fig:ir-share}), which implicitly reflects how aquifer thickness alters well yields and, in turn, farmers' ability to effectively and reliably buffer crops against water deficits during the growing season.

@@

{\centering \includegraphics[width=6in,height=500px,]{../../Figures/g_total_impact}

@@

\caption{Average productivity of corn and soybean in the US High Plains for different levels of water deficit and aquifer thickness, considering both per-area yield responses and farmers' choices about irrigated area shares in response to different water deficit and aquifer conditions. See Appendix \ref{reg-conf} for estimated average yield response curves with 95\% confidence intervals displayed, which are omitted here for visual clarity.}\label{fig:tot-impact}
\end{figure}

@@

{\centering \includegraphics[width=6in,height=500px,]{../../Figures/g_avg_yield_dif}

@@

\caption{Difference in average yield of corn and soybean for aquifer thickness levels of 40 m, 70 m, and 100 m, relative to a baseline aquifer thickness of 10 m. Shaded areas denote 95\% confidence interval, which when greater than zero indicate average yields for the given aquifer thickness level are statistically significantly greater than for an aquifer thickness of 10 m at the 5\% level of significance.}\label{fig:dif-tot-impact}
\end{figure}

@@

Second, our analysis considered the effects of aquifer thickness on farmers' decisions to switch from irrigated to rainfed production, but did not assess other potential responses to aquifer depletion such as changing crop types or irrigation management practices. Switching to more drought resistant crops or varieties may enable farmers to maintain irrigated production at lower levels of aquifer thickness, and has been shown to be an important response to increasing agricultural water scarcity and abstraction policies in other studies \citep{bhattarai2021impact, deines2019quantifying, manning2017producer}. Improvements to irrigation scheduling practices and technologies, such as have been implemented as part of the SD6-LEMA (Local Enhanced Management Area) in Kansas \citep{deines2019quantifying, glose2022quantifying} and in parts of Texas \citep{mrad2020peak} and which are proposed as part of California's Sustainable Groundwater Management Act (SGMA) \citep{berbel2019droughts, lubell2020sustainable}, could also help to improve the efficiency of irrigation and, hence, the ability to more effectively meet crop water demands with lower well capacities. Such interventions may reduce - but not eliminate - increases in drought risk caused by lower or declining aquifer thickness, and could also help to mitigate against potential yield losses caused by over-use of water in wetter years via raising awareness of efficiency saving potentials \citep{foster2019assessing} and the importance of conservation \citep{marston2022}. However, risks also exist that irrigation efficiency improvements will increase net extraction, ultimately causing a rebound effect that exacerbates future aquifer depletion and drought risks \citep{grafton2018paradox,perez2021agricultural}. Future research should explore potential strategies that farmers could adopt to lessen the combined impacts of drought and aquifer depletion, alongside efforts to curb unsustainable groundwater extraction and climate change.

Finally, we focus on the impacts of aquifer thickness on agricultural resilience to seasonal water deficits. However, where water deficits are distributed unevenly during the season, negative impacts of aquifer depletion may be even larger. This is because reductions in well yields caused by aquifer depletion will constrain farmers' ability to meet peak crop water demands, increasing yield losses compared with if water deficits had been distributed evenly throughout the season \citep{ortiz2019unpacking}. Furthermore, irrigation also has broader benefits for crops such as reducing heat-related stress through cooling of land surface temperatures \citep{adegoke2003impact, bonfils2007empirical, lobell2008effect, zhu2022untangling}. Reductions in aquifer thickness therefore may have other negative impacts on irrigated crop productivity, if constraints to water supply due to aquifer depletion are significant enough to limit cooling benefits of irrigation during the growing season.
